---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## My first R Markdown report: a new glance of reproducible research

This is a report of a exploratory analysis from a dataset of activity monitoring data, recorded from devices such as Fitbit or Nike Fuelband.   

For each observation, the dataset contains three (3) variables:  

* steps: Number of steps in a 5-minute interval.  
* date: The date on which the measurement was taken.  
* interval: an identifier for the 5-minute interval in which the measurement was taken.  


### Loading the dataset

The dataset was downloaded in a csv format and loaded into R with the ```readr()``` package
in order to obtain a tibble to work on with *tydiverse* packages. Dates columns were formated as such.

```{r}
library(readr)
activity <- read_csv("activity.csv", col_types = cols(date = col_date(format = "%Y-%m-%d"), 
                                                      steps = col_double()), na = "NA") # Read data file.
```

### Mean total number of steps per day (*First task*)

In order to evaluate the mean number of steps per day, some transformations were necesary in the dataset. Rows with ```NA``` were eliminated for this analysis and the dates were treated as a ```factor()``` in order to compute the sum of steps taken each day. 

```{r message=FALSE}
library(dplyr)
```
```{r}
activity_wo_na <- na.omit(activity) # Eliminate rows with NAs

activity_wo_na$interval <- as.factor(activity_wo_na$interval) # Transform interval in a Factor variable

activity_w_dates <- mutate(activity_wo_na, date_factor = as.factor(date)) %>%
                        group_by(date_factor) # Create a new date factor and group the data by this factor

head(activity_w_dates)
```

The ```sumarise()``` function was used to obtain the total steps per day (with the ```sum()``` function), the mean and the median steps per day. The ```date_factor``` variable was a used as a grouping variable.

```{r}
steps_by_day <- summarise(activity_w_dates, step_number = sum(steps),
                          steps_mean = mean(steps), steps_median = median(steps))
print(steps_by_day)
```

The results stored in this new ```tibble``` were used to generate the histogram for the total number of steps per day and the mean number of steps per day. This final result will be presented in a barplot.

```{r}
library(ggplot2)
hist <- ggplot(data = steps_by_day, aes(steps_by_day$step_number)) + 
        geom_histogram(binwidth = 2500, col = "black", fill = "lightblue") +
        labs(title = "Total number of steps by day", x = "Number of steps",
             y = "Count")

n.df <- data.frame(days = as.character(steps_by_day$date_factor), 
                   stepsmean = steps_by_day$steps_mean)

bars <- ggplot(data = n.df, aes(x = days, y = stepsmean)) +
        geom_bar(stat = "identity", col = "black", fill = "lightblue") + 
        theme(axis.text.x = element_text(angle=60, hjust=1)) +
        labs(title = "Mean number of steps by day", x = "Day",
        y = "Mean number of steps")
                       
print(hist)
print(bars)
```

### Average daily activity pattern (*Second task*)

For this section of the analysis, the `activity_wo_na` ```tible``` (previously created) was used. To analyze the daily activity pattern, `interval` was used as grouping variable. 

```{r}
activity_w_intervals <- group_by(activity_wo_na, interval)

print(activity_w_intervals)
```

The average steps by interval (for all days) was calculated as follows

```{r}
steps_by_interval <- summarise(activity_w_intervals, Avrg.Steps = mean(steps))

print(steps_by_interval)

time_plot_1 <- ggplot(data = steps_by_interval, (aes(x = as.numeric(levels(interval)), 
                                                     y = Avrg.Steps))) +
               geom_line() + labs(title = "Time series: 5 minute interval vs. Average
                                  number of steps taken",
                                  x = "5 minute interval",
                                  y = "Average number of steps (all days)")
               
print(time_plot_1)
```

On average, the interval `835` contains the maximum number of steps

```{r}
max_interval <- steps_by_interval[which.max(steps_by_interval$Avrg.Steps),]

print(max_interval)
```

### Working with missing values (`NA`). Inputing them and evaluating its effects on the data (*Third task*)

We start again loading the "main" dataset: the raw data with no transformations.

```{r}
activity <- read_csv("activity.csv", col_types = cols(date = col_date(format = "%Y-%m-%d"), 
                                                      steps = col_double()), na = "NA") # Read data file.
```

The number of `NA`s in the dataset were count by column, to evaluate if these missing values were present in just one or other columns

```{r}
NAs_per_var <- apply(X = activity, MARGIN = 2, FUN = function(X){sum(is.na(X))})

print(NAs_per_var)
```

The `interval` variable was transformed `as.factor()` bo used to `left_join()` the `activity` dataset with the previously created `steps_by_interval` dataset. This final one contains the average steps by 5-minute interval. In this sense, the `NA`s in the `activity` dataset were substituted for the mean steps of the 5-minute interval representative of a particular missing value. 

```{r}
activity$interval <- as.factor(activity$interval)

activity_w_mean_steps <- left_join(x = activity, y = steps_by_interval, by = "interval")

head(activity_w_mean_steps)

cases <- which(is.na(activity_w_mean_steps$steps) == TRUE)

activity_w_mean_steps$steps[cases] <- activity_w_mean_steps$Avrg.Steps[cases]

data_filled_NA <- activity_w_mean_steps
head(data_filled_NA)
```

The mean, median and total number of steps taken by day were calculated. An histogram is plotted to compare these results with the ones of the first task (where the `NA`s were omited)

```{r}
act_filled_NA <- mutate(data_filled_NA, date_factor = as.factor(date)) %>% group_by(date)

steps_by_day_filled <- summarise(act_filled_NA, step_number = sum(steps),
                          steps_mean = mean(steps), steps_median = median(steps))

head(steps_by_day_filled)

hist_2 <- ggplot(data = steps_by_day_filled, aes(steps_by_day_filled$step_number)) + 
          geom_histogram(binwidth = 2500, col = "black", fill = "lightblue") +
          labs(title = "Total number of steps by day (NAs replaced)", x = "Number of steps",
          y = "Count")

print(hist_2)
print(hist)
```

When comparing the histograms, it could be noticed that **the count is increased while filling the `NA`s with new data**, as expected because now there is no omited rows. Nevertheless, **the behavior of the variable is still very similar**, because the mean values that filled the `NA`s are representatives of the tendency of the values taken by the `steps` variable for each 5-minute interval. 

### Evaluating differences between activity patterns between weekdays and weekends (*Fourth task*)

A new factor variable `is.weekday` was created to differenciate between weekends and weekdays in the dataset with the final objective of making comparissons between these conditions

```{r}
data_filled_NA

w_days <- mutate(data_filled_NA, day = weekdays(date)) %>%
          mutate(is.weekday = factor((day %in% c("lunes", "martes","miércoles","jueves","viernes")), levels = c(FALSE,TRUE), 
                                      labels = c("Weekend", "Weekday")))

w_days_g <- group_by(w_days, interval, is.weekday)

head(w_days_g)

w_days_sg <- summarise(w_days_g, Avrg.Steps = mean(steps))

time_plot_final <- ggplot(data = w_days_sg, (aes(x = as.numeric(interval), y = Avrg.Steps))) +
                   geom_line() + labs(title = "Time series: 5 minute interval vs. Average number of steps taken",
                                  x = "5 minute interval",
                                  y = "Average number of steps (all days)") + 
                   facet_grid(is.weekday ~.) + 
                   scale_x_continuous(breaks = c(0,57.6,115.2,172.8,230.4), 
                                      labels = c("0","500","1000","1500","2000"))
                   
print(time_plot_final)
```